\thetitle

### 2021-03-03 writing/discussion.tex
claeify language of threats, push experiment

\subsection{Threat: Is the Rational Programmer Realistic?}

@@

\subsection{Threat: Is our Transient Correct?}
\label{sec:threat:transient}

As we discuss in section~\ref{sec:results}, Transient surprisingly
produces
an empty blame set in 967 scenarios.
Such cases should never occur, at least in theory, because an empty set
 means the value has never crossed a boundary---if the value is indeed defined
 in the current module, then we have typed code blaming itself for a typed
 value as all the checks are inlined in typed code. Thus, the type checker must be unsound.
After careful investigation of these empty blame cases, we found no soundness
 bugs in Transient Typed Racket.
Instead, we found scenarios in which Transient lost track of the proper
 boundaries.
These scenarios suggest ways to improve~\citet{vss-popl-2017}'s blame
algorithm:
\begin{itemize}
  \item
    Entries in a blame map must point to several parent entries.
    For example, if the function \texttt{f} receives bad input in the call
    \texttt{(filter f xs)}, then blame should point to both \texttt{filter}
    and the \texttt{xs} list.
  \item
    The construction of blame map entries must be guided by type-like specifications
     instead of relying on syntax.
    Aliasing the built-in \texttt{filter} function should not change the shape
     of the blame map.
  \item
    The initial blame map must reflect the initial type environment.
    For instance, Typed Racket trusts that untyped core-library functions, such as \texttt{filter},
     behave correctly; such assumptions should appear in the blame map.
\end{itemize}
\noindent{}Despite these known limitations, based on the relatively small
number of debugging failures, we conjecture that these improvements
 to Transient blame-tracking do not affect our overall conclusions.

A second apparent flaw in our Transient implementation is the high cost of blame.
Whereas \citet{vss-popl-2017} report an average slowdown of 2.5x and
worst-case of 5.4x on fully-typed benchmarks in Reticulated,
some of our configurations exceed our 4 minute timeout or 6GB memory limit with transient blame.
To put those limits into context, the fully typed and fully untyped benchmarks all complete in a few seconds with minimal memory usage, and none of the mixed Natural configurations hit these limits.
Unfortunately, our high cost for Transient seems closer to the truth;
 there are at least three broad issues that skew the earlier results.
First, the 2017 implementation of Reticulated fails to insert certain
 soundness checks\footnote{Missing check: \url{https://github.com/mvitousek/reticulated/issues/36}}
 and blame-map updates\footnote{Missing cast: \url{https://github.com/mvitousek/reticulated/issues/43}}
 from the paper.
Second, Reticulated's type checker assigns the dynamic type to local
 variables in the "fully typed" benchmarks---often because the type system
 cannot articulate a more precise type.
Code that ends up with the dynamic type has fewer constraints to check at run-time.
Third, \citet{vss-popl-2017} use relatively small benchmarks.
Four have since been retired from the official Python benchmark suite
 because they are too small, unrealistic, and unstable.\footnote{Release notes: \url{https://pyperformance.readthedocs.io/changelog.html}}
On larger programs, Reticulated suffers from high overhead.
For example, a Reticulated version of our simplest benchmark runs in
 ~40 seconds without blame, and with blame times out after 10 minutes.

>>>>

%% -----------------------------------------------------------------------------
\subsection{Threat: Is the Rational Programmer Realistic?}

@@

%% -----------------------------------------------------------------------------
\subsection{Threat: Can the Experiment Improve Transient?} \label{sec:threat:transient}

The results in section~\ref{sec:results} show that Transient produces empty
blame sequences for 967 scenarios. In theory, empty sequence must not occur. An
empty blame sequence means a lack of boundary crossings for the witness value.
By implication, a typed module is blaming itself---something that can
happen only if the type checker (or system) is unsound.

An investigation of these empty blame cases reveals that Vitousek et al.'s
Transient blame assignment can lose track of the proper boundaries.  Despite
these losses, the number of failures is a small fraction of the overall number
of debugging scenarios. It is unlikely that any improvements to Vitousek et
al.'s strategy would change the overall conclusion about Transient blame.

{\bf Note} These scenarios suggest ways to improve Transient's blame map as
follows:
\begin{itemize}

\item Entries in a blame map should point to {\em several\/} parent entries.
 For example, if \texttt{f} receives bad input in the context of {\tt (filter f
 xs)}, the entry should point to both \texttt{filter} and \texttt{xs}.

\item The construction of entries must be guided by type-like specifications for
 primitives instead of special cases for built-in functions. Aliasing such
 function changes the blame map.

%% I don't see what this really means or contributes to the first two bullets:

 % \item The initial blame map must reflect the initial type environment.  For
 % instance, Typed Racket trusts that untyped core-library functions behave
 % correctly.

\end{itemize}
This small example is indicative of how rational-programmer experiments may
provide feedback.

%% -----------------------------------------------------------------------------
\subsection{Threat: Is the Transient Blame Assignment Mechanism Realistic?}
\label{sec:threat:transient2}

The results in section~\ref{sec:results} also show that the cost of Transient
blame is quite high. Under the Transient semantics, some of the debugging
scenarios exceed the 4-minute timeout or the 6GB-memory limit. To put those
limits into context, the fully typed and fully untyped benchmarks all complete
in a few seconds with minimal memory usage. Furthermore, none of the mixed
Natural configurations hit these limits, and with the blame map turned off,
the Transient semantics also runs these programs in a short amount of time and
well within the memory limits. In short, even though the Transient rational
programmer appears to do well in the experiment, the implementation of the
Transient blame strategy might be unrealistic.

At first glance, these measurements seem to contradict the results
of~\citet{vss-popl-2017}. They report an average slowdown of 2.5x and a
worst-case of 5.4x on fully-typed benchmarks in Reticulated Python.
Unfortunately, the high cost of Transient Racket seems closer to the truth.
There are at least four broad issues that skew the earlier results:
\begin{enumerate}

\item The 2017 implementation of Reticulated fails to insert certain soundness
checks\footnote{Missing check:
\url{https://github.com/mvitousek/reticulated/issues/36}} and blame-map
updates\footnote{Missing cast:
\url{https://github.com/mvitousek/reticulated/issues/43}} from the paper.

\item The types of almost all local variables in the "fully typed" benchmarks is
{\tt Dynamic}, often due the impoverished nature of the type system, which does
not allow the ascription of precise types.  Code that ends up with the type {\tt
Dynamic} has fewer constraints to check at run-time---and much less information
to track in the blame map.

\item \citet{vss-popl-2017} use small benchmarks.  Four have since been retired
from the official Python benchmark suite because they are too small,
unrealistic, and unstable.\footnote{Release notes:
\url{https://pyperformance.readthedocs.io/changelog.html}}

\item On large programs, Reticulated suffers from high overhead.  For example, a
Reticulated version of the simplest benchmark runs in ~40 seconds without blame,
and with blame times out after 10 minutes. Typed Racket runs the same benchmark
in a small fraction of the time.
\end{enumerate}
Future work on Transient blame is needed to make an informed decision. The
evaluation process presented here can help guide this research.
